---
title: "Gapminder exploration"
author: "Amanda Cheung"
date: "9/26/2017"
output: github_document
---
*Exploring the Gapminder data for STAT 545 homework 02*

### Bring rectangular data in

Load gapminder and dplyr via the tidyverse meta-package.
```{r}
library(gapminder)
library(tidyverse)
```

### Smell test the data

1. Is it a data.frame, a matrix, a vector, a list? What's its class?

*Display the structure of the gapminder object.*
```{r}
str(gapminder)
```
*Alternatively, you can display the class.*
```{r}
class(gapminder)
```

Gapminder is a data.frame. It's class tbl_df, tbl, and data.frame.

2. How many variables/columns? How many rows/observations?
*Find the number of columns and rows.*
```{r}
ncol(gapminder)
nrow(gapminder)
```
Gapminder has 6 variables/columns and 1604 rows/observations. This information is also seen above with str().

3. Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?
Alternative ways to get facts about "extent" or "size".
```{r}
summary(gapminder)
dim(gapminder)
length(gapminder)
```

Different functions can be used depending on the type of information you want.
str() provides an overview of the object. 
class() provides information about the class of the object.
ncol() and length() provides the number of columns/length of the object.
nrow() provides the number of rows of the object.
dim() provides the dimensions, number of rows x number of columns, of the object.
summary() provides a statistical overview.

4. What data type is each variable?
From the structure function, str(), above:

variable  | data type
----------|----------
country   | factor
continent | factor
year      | integer 
lifeExp   | numeric
pop       | integer
gdpPercap | numeric










