---
title: "Gapminder exploration"
author: "Amanda Cheung"
date: "9/26/2017"
output: github_document
---
Exploring the Gapminder data for STAT 545 homework 02

## Bring rectangular data in

Load gapminder and dplyr via the tidyverse meta-package.
```{r}
library(gapminder)
library(tidyverse)
```

## Smell test the data

**Is it a data.frame, a matrix, a vector, a list? What's its class?**

Display the structure of the gapminder object.
```{r}
str(gapminder)
```
Alternatively, you can display the class.
```{r}
class(gapminder)
```

- Gapminder is a data.frame. It is class tbl_df, tbl, and data.frame.

**How many variables/columns? How many rows/observations?**

Find the number of columns and rows.
```{r}
ncol(gapminder)
nrow(gapminder)
```
- Gapminder has 6 variables/columns and 1604 rows/observations. This information is also seen above with str().

**Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**

- Alternative ways to get facts about "extent" or "size".

```{r}
summary(gapminder)
dim(gapminder)
length(gapminder)
```

- Different functions can be used depending on the type of information you want.
    + str() provides an overview of the object. 
    + class() provides information about the class of the object.
    + ncol() and length() provides the number of columns/length of the object.
    + nrow() provides the number of rows of the object.
    + dim() provides the dimensions, number of rows x number of columns, of the object.
    + summary() provides a statistical overview.

**What data type is each variable?**

From the structure function, str(), above:

variable  | data type
----------|----------
country   | factor
continent | factor
year      | integer 
lifeExp   | numeric
pop       | integer
gdpPercap | numeric

## Explore individual variables

### Continent as a categorical variable

**What are possible values (or range, whichever is appropriate) of each variable?**

Looking at the summary, table, and levels of the continent variable.
```{r}
summary(gapminder$continent)
table(gapminder$continent)
levels(gapminder$continent)
```
- The continent variable has 5 values.
1. Africa
2. Americas
3. Asia
4. Europe
5. Oceania

**What values are typical? What’s the spread? What’s the distribution?**

Observations associated with each continent in a bar graph.
```{r}
ggplot(gapminder, aes(x=continent, fill=continent)) + geom_bar()
```

Observations associated with each continent in a density plot.
```{r}
ggplot(gapminder, aes(x=continent, fill=continent, colour=continent)) + geom_density(alpha=0.5)
```

Observations associated with each continent in a pie chart.
```{r}
pie(table(gapminder$continent), col=c("blue","red","yellow", "orange", "green"), main="Observations associated with each continent")
```

### Life expectancy as a quantitative variable

**What are possible values (or range, whichever is appropriate) of each variable?**

Looking at the summary, head and tail values, and range of the life expectancy variable.
```{r}
summary(gapminder$lifeExp)
head(gapminder$lifeExp)
tail(gapminder$lifeExp)
range(gapminder$lifeExp)
```

**What values are typical? What’s the spread? What’s the distribution?**

Histogram of life expectancy values to illustrate the spread and distribution.
```{r}
ggplot(gapminder, aes(x=lifeExp, fill=continent)) + geom_histogram(binwidth=10)
```

Density plot of life expectancy values to illustrate the spread and distribution.
```{r}
ggplot(gapminder, aes(x=lifeExp)) + geom_density(colour="blue", fill="blue", alpha=0.25)
```

## Explore various plot types

#### Scatterplot

Scatterplot of year vs. life expectancy in Europe.
```{r}
ggplot(filter(gapminder, continent == "Europe" ), aes(x=year, y=lifeExp, colour=country)) + geom_line() + geom_point()
```

Scatterplot of population vs. life expectancy in the '80s and '90s in Africa, the Americas, and Asia.
```{r}
ggplot(gapminder %>% 
         filter(year <= 1999, year >= 1980, continent %in% c("Americas", "Africa", "Asia")) %>% 
         select(continent, pop, lifeExp, gdpPercap),
         aes(x=pop, y=lifeExp, colour=continent)) + geom_point(aes(size=gdpPercap), alpha=0.5) + geom_smooth(se=FALSE, method="loess") + scale_x_log10()
```

```{r}
ggplot(gapminder, aes(x=gdpPercap, y=pop)) + scale_y_log10() + facet_wrap(~ continent) + geom_point(alpha=0.25) + geom_smooth(se=FALSE, method="loess", aes(colour=continent))
```

#### Frequency polygon

Frequency polygon of life expectancy in China, Japan, Vietnam, and Thailand.
```{r}
ggplot(gapminder %>%  
         filter(continent == "Asia", country %in% c("China", "Japan", "Vietnam", "Thailand")) %>%
         select(country, lifeExp), 
         aes(x=lifeExp, colour=country)) + geom_freqpoly(binwidth=10)
```

#### Histogram

Histogram of GDP in billions (calculated by multiplying GPD per capita with population) in Oceania.
```{r}
ggplot(gapminder %>% 
         mutate(gdpBill = (gdpPercap * pop)/1000000000) %>% 
         filter(continent == "Oceania") %>% 
         select(gdpBill, country), 
         aes(x=gdpBill, fill=country)) + geom_histogram(binwidth=50)
```

#### Boxplot

The distribution of population of each continent in the '90s shown in a boxplot. 
```{r}
ggplot(gapminder %>% 
         filter(year <= 1999, year >= 1990),
       aes(x=continent, y=pop, colour=continent)) + 
  geom_boxplot(outlier.alpha = 0.5, outlier.shape = 1) + scale_y_log10()
```

#### Area chart

The spread of GPD per capita each year in Algeria, Canada, Denmark, and India.   
```{r}
ggplot(gapminder %>% 
         filter(country %in% c("Canada", "India", "Denmark", "Algeria")),
       aes(x=year, y=gdpPercap, fill=country)) + geom_area(colour="black", alpha=0.6)
```

## I want to do more!

```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```

The code above does not suceed in retrieving all the data for Afghanistan and Rwanda. The "==" operator does not work for more than one category where there is overlapping data (i.e. if Afghanistan and Rwanda share the same year, only one set of data will be retrieved).

The "%in%" operator is used along with c(), to create a vector, allowing retrieval of all the data where both countries share the same element (i.e. year). 

```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```