---
title: "Gapminder exploration"
author: "Amanda Cheung"
date: "9/26/2017"
output: github_document
---
Exploring the Gapminder data for STAT 545 homework 02

## Bring rectangular data in

Load gapminder and dplyr via the tidyverse meta-package.
```{r}
library(gapminder)
library(tidyverse)
```

## Smell test the data

**Is it a data.frame, a matrix, a vector, a list? What's its class?**

Display the structure of the gapminder object.
```{r}
str(gapminder)
```
Alternatively, you can display the class.
```{r}
class(gapminder)
```

- Gapminder is a data.frame. It is class tbl_df, tbl, and data.frame.

**How many variables/columns? How many rows/observations?**

Find the number of columns and rows.
```{r}
ncol(gapminder)
nrow(gapminder)
```
- Gapminder has 6 variables/columns and 1604 rows/observations. This information is also seen above with str().

**Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in different contexts?**

- Alternative ways to get facts about "extent" or "size".

```{r}
summary(gapminder)
dim(gapminder)
length(gapminder)
```

- Different functions can be used depending on the type of information you want.
    + str() provides an overview of the object. 
    + class() provides information about the class of the object.
    + ncol() and length() provides the number of columns/length of the object.
    + nrow() provides the number of rows of the object.
    + dim() provides the dimensions, number of rows x number of columns, of the object.
    + summary() provides a statistical overview.

**What data type is each variable?**

From the structure function, str(), above:

variable  | data type
----------|----------
country   | factor
continent | factor
year      | integer 
lifeExp   | numeric
pop       | integer
gdpPercap | numeric

## Explore individual variables

### Continent as a categorical variable

**What are possible values (or range, whichever is appropriate) of each variable?**

Looking at the summary, table, and levels of the continent variable.
```{r}
summary(gapminder$continent)
table(gapminder$continent)
levels(gapminder$continent)
```
- The continent variable has 5 values.
1. Africa
2. Americas
3. Asia
4. Europe
5. Oceania

**What values are typical? What’s the spread? What’s the distribution?**

Observations associated with each continent in a bar graph.
```{r}
ggplot(gapminder, aes(x=continent, fill=continent)) + geom_bar()
```

Observations associated with each continent in a density plot.
```{r}
ggplot(gapminder, aes(x=continent, fill=continent, colour=continent)) + geom_density(alpha=0.5)
```

Observations associated with each continent in a pie chart.
```{r}
pie(table(gapminder$continent), col=c("blue","red","yellow", "orange", "green"), main="Observations associated with each continent")
```













