---
title: "STAT545 Homework 05: Factor and Figure Management"
output: github_document
---

The final homework assignment for STAT545 (until STAT547...) -- let's do this! :neckbeard:

### Load packages

Load tidyverse, forcats, and knitr.

```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(forcats))
suppressPackageStartupMessages(library(knitr))
```

### Locate Gapminder data

```{r}
## Locate gapminder package on my system
(gap_tsv <- system.file("gapminder.tsv", package = "gapminder"))
```


## Factor Management

### GAPMINDER VERSION

### 1. Factorise

Import gapminder data using `read_tsv()`, which reads tab-delimited data. 

```{r}
gapminder <- read_tsv(gap_tsv)
str(gapminder, give.attr = FALSE)
```

Strings are not converted to factors, so let's convert `country` and `continent` to factors.

```{r}
gapminder <- gapminder %>%
  mutate(country = factor(country),
         continent = factor(continent))
## as_factor() could also be used in lieu of factor()
## as_factor() is actually safer, since it doesn't change the order of the levels
## however, not important right now since we'll be reordering them later!

str(gapminder)
```

Ta da! We now see that `country` is a factor with 142 levels, and `continent` is a factor with 5 levels.

### 2. Drop Oceania

Filter the Gapminder data to remove observations associated with the `continent` of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors.

#### Before removing Oceania and unused factor levels...

```{r}
str(gapminder) ## structure of original gapminder
nrow(gapminder) ## total number of rows of original gapminder
levels(gapminder$continent) ## levels in original gapminder continent
nlevels(gapminder$continent) ## total number of levels in original gapminder continent
class(gapminder$continent) ## class of original gapminder continent
summary(gapminder$continent) ## summary of original gapminder continent
```

#### Drop Oceania and unused factor levels. 

```{r}
## new_gap will be the filtered gapminder data frame
new_gap <- gapminder %>%
  filter(continent != "Oceania") %>% 
  droplevels()
```

#### After removing Oceania and unused factor levels...

```{r}
str(new_gap) ## structure of filtered gapminder
nrow(new_gap) ## total number of rows of filtered gapminder
levels(new_gap$continent) ## levels in filtered gapminder continent
nlevels(new_gap$continent) ## total number of levels of filtered gapminder continent
class(new_gap$continent) ## class of filtered gapminder continent
summary(new_gap$continent) ## summary of filtered gapminder continent
```

After removing Oceania and unused factor levels, the number of rows decreased to 1680 (from 1704), there are 4 levels in continent, and the levels no longer contain Oceania.

### 3. Reorder the levels of `country` or `continent`

Use the `forcats` package to change the order of the factor levels, based on a principled summary of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the median.

Order gapminder countries by maximum population in descending order. This will display the top 10 countries with the greatest maximum population. 

```{r warning=FALSE}
fct_reorder(gapminder$country, gapminder$pop, max, .desc = TRUE) %>% 
  levels() %>%
  head(10) %>% 
  kable()
```

Order gapminder countries by maximum population. This will display the top 10 countries with the smallest maximum population.

```{r warning=FALSE}
fct_reorder(gapminder$country, gapminder$pop, max) %>% 
  levels() %>%
  head(10) %>%
  kable()
```

Let's look at the importance of reordering factor levels! Here is a plot of gapminder European countries in 2002 to population. 

```{r}
gap_euro <- gapminder %>% filter(year == 2002, continent == "Europe")
ggplot(gap_euro, aes(x = country, y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

From this plot, we can see that Germany has the greatest population, but it's hard to see the top 10 countries.  

```{r}
ggplot(gap_euro, aes(x = fct_reorder(country, pop), y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

Ahhh, much better! Now we can easily differentiate the countries with the greatest population.

### `arrange()`

#### Explore the effects of `arrange()`. Does merely arranging the data have any effect on, say, a figure?

Arrange population in `gap_euro` which was created above (gapminder European countries in 2002). 
```{r}
gap_euro2 <- gapminder %>% 
  filter(year == 2002, continent == "Europe") %>% 
  arrange(pop)
```

#### Let's see how the **data** has changed before and after `arrange()`.

**Before `arrange()`**
```{r warning=FALSE}
## row order
gap_euro %>% head() %>% kable() 

## level order
gap_euro$country %>% fct_drop() %>% levels() %>% head() %>% kable()
```

**After `arrange()`**
```{r warning=FALSE}
## row order
gap_euro2 %>% head() %>% kable()

## level order
gap_euro2$country %>% fct_drop() %>% levels() %>% head() %>% kable()
```

Before `arrange()`, countries were listed in alphabetical order in the row order and level order of the data frame.

After `arrange()`, countries were listed in increasing order of maximum population in the row order, but the level order remained alphabetical.   

#### Let's see how the **figure** has changed before and after `arrange()`.

**Before `arrange()`**
```{r}
ggplot(gap_euro, aes(x = country, y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

**After `arrange()`**
```{r}
ggplot(gap_euro2, aes(x = country, y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

No change! Arranging the data does not have any effect on a figure. However, it changes the row order in a data frame. 

#### Explore the effects of reordering a factor and factor reordering coupled with `arrange()`. Especially, what effect does this have on a figure?

#### Let's see how the data has changed before and after `arrange()`.

**Before `arrange()`**
```{r warning=FALSE}
gap_euro_reorder <- gapminder %>% 
  filter(year == 2002, continent == "Europe") %>% 
  mutate(country = fct_reorder(country, pop))

## row order
gap_euro_reorder %>% head() %>% kable()

## level order
gap_euro_reorder$country %>% fct_drop() %>% levels() %>% head() %>% kable()
```


**After `arrange()`**
```{r warning=FALSE}
gap_euro_reorder2 <- gapminder %>% 
  filter(year == 2002, continent == "Europe") %>% 
  mutate(country = fct_reorder(country, pop)) %>% 
  arrange(pop)

## row order
gap_euro_reorder2 %>% head() %>% kable()

## level order
gap_euro_reorder2$country %>% fct_drop() %>% levels() %>% head() %>% kable()
```

Before `arrange()`, the row order of countries were listed in alphabetical order. Level order was changed to countries listed in increasing order of maximum population.

After `arrange()`, the row order and level order were changed to countries listed in increasing order of maximum population. 

#### Let's see how the **figure** has changed before and after `arrange()`.

**Before `arrange()`**
```{r}
ggplot(gap_euro_reorder, aes(x = country, y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

**After `arrange()`**
```{r}
ggplot(gap_euro_reorder2, aes(x = country, y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```

No change! Arranging the data does not have any effect on a figure that has reordered factors. However, it changes the row order in a data frame.

From this activity, we can see that `arrange()` does not effect the order of levels in a factor, but instead it changes the order of rows in a data frame.

## File I/O

Experiment with one or more of `write_csv()/read_csv()` (and/or TSV friends), `saveRDS()/readRDS()`, `dput()/dget()`. Create something new, probably by filtering or grouped-summarization of Gapminder. Fiddle with the factor levels, i.e. make them non-alphabetical (see previous section). Explore whether this survives the round trip of writing to file then reading back in.

### Create something new!

Let's make a country level summary of maximum population.
```{r}
gap_max_pop <- gapminder %>%
  filter(year == 2002) %>% 
  group_by(country, continent) %>% 
  summarise(max_pop = max(pop)) %>% 
  ungroup()
kable(head(gap_max_pop))
```

### Reorder the data!

```{r warning=FALSE}
## countries are in alphabetical order
kable(head(levels(gap_max_pop$country)))

## reorder to increasing maximum population
gap_max_pop <- gap_max_pop %>% 
  mutate(country = fct_reorder(country, max_pop))
kable(head(levels(gap_max_pop$country)))

## the row order has not changed
## but the level order changed!
kable(head(gap_max_pop))
```


### Write the data out!

```{r}
write_csv(gap_max_pop, "gap_max_pop.csv") ## write to comma-delimited file
saveRDS(gap_max_pop, "gap_max_pop.rds") ## save to binary file
dput(gap_max_pop, "gap_max_pop-dput.txt") ## write to plain text
```


### Read the data!

```{r}
gap_max_pop_csv <- read_csv("gap_max_pop.csv")
gap_max_pop_RDS <- readRDS("gap_max_pop.rds")
gap_max_pop_dget <- dget("gap_max_pop-dput.txt")
```

### Compare the different strategies!

```{r}
## first convert countries to factors
gap_max_pop_csv <- gap_max_pop_csv %>% mutate(country = factor(country))
gap_max_pop_RDS <- gap_max_pop_RDS %>% mutate(country = factor(country))
gap_max_pop_dget <- gap_max_pop_dget %>% mutate(country = factor(country))

## create a tibble to compare the strategies
country_levels <- tibble(original = head(levels(gap_max_pop$country)))
country_levels <- country_levels %>%
  mutate(via_csv = head(levels(gap_max_pop_csv$country)),
         via_RDS = head(levels(gap_max_pop_RDS$country)),
         via_dput = head(levels(gap_max_pop_dget$country)))
kable(country_levels)
```

The original (after reordering) country factor levels are restored using the `saveRDS() / readRDS()` and `dput() / dget()` strategy, but return to alphabetical ordering using `write_csv() / read_csv()`.

## Visualization design

Remake at least one figure or create a new one, in light of something you learned in the recent class meetings about visualization design and color. Maybe juxtapose your first attempt and what you obtained after some time spent working on it. Reflect on the differences. If using Gapminder, you can use the country or continent color scheme that ships with Gapminder. Consult the guest lecture from Tamara Munzner and everything here.

### Remake plot of European countries to maximum population in 2002 (from above)!

```{r}
ggplot(gap_euro, aes(x = fct_reorder(country, pop), y = pop)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw()
```








## Clean up your repo!

## But I want to do more!

```{r}
country <- c('Argentina', 'Austria', 'Bangladesh', 'Chad', 
             'Ecuador', 'Gambia', 'Italy', 'Nepal', 
             'New Zealand', 'Switzerland', 'Taiwan', 'United Kingdom')
capital_city <- c('Buenos Aires', 'Vienna', 'Dhaka', "N'Djamena", 
                  'Quito', 'Banjul', 'Rome', 'Kathmandu',
                  'Wellington', 'Bern', 'Taipei', 'London')

country_data <- data.frame(country, capital_city)

kable(country_data)
```





