---
title: "Dplyr and ggplot2 exploration"
author: "Amanda Cheung"
date: "10/1/2017"
output: github_document
---

## STAT545 Homework 03

Exploration and manipulation of the Gapminder dataset using dplyr and ggplot2.

## Gapminder, dplyr, and ggplot2
Load gapminder, and dplyr and ggplot2 via the tidyverse meta-package.
```{r}
library(gapminder)
library(tidyverse)
```

## Task 1: Get the maximum and minimum of GDP per capita for all continents

#### Table 1: Maximum and minimum GDP per capita for all continents.
```{r results= 'asis'}
knitr::kable(gapminder %>%
               group_by(continent) %>% 
               summarize(maximum_gdpPercap = max(gdpPercap),
                         minimum_gdpPercap = min(gdpPercap)))
```

#### Figure 1: Grouped bar plot of maximum and minimum GDP per capita for all continents.

Load reshape2 package.
```{r}
library(reshape2)
```
Melt data of interest using `melt` function.
```{r}
gdp_melt = melt(gapminder %>%
       group_by(continent) %>% 
       summarize(maximum_gdpPercap = max(gdpPercap),
                 minimum_gdpPercap = min(gdpPercap)))
```

Grouped bar plot of maximum and minimum GDP per capita for all continents
```{r}
ggplot(gdp_melt, aes(x=continent, y=value, fill=variable)) +
  geom_bar(position="dodge", stat="identity") + 
  theme_bw() +
  theme(legend.position="top", legend.title = element_blank(),
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(labels=c("Maximum", "Minimum")) +
  labs(x="Continent", y="GDP per capita") +
  ggtitle("Maximum and minimum GDP per capita for all continents")
```

## Task 2: Look at the spread of GDP per capita within the continents.

#### Table 1: Maximum, minimum, median, average, and standard deviation of GDP per capita for all continents.
```{r results= 'asis'}
knitr::kable(gapminder %>%
               group_by(continent) %>% 
               summarize(maximum_gdpPercap = max(gdpPercap),
                         minimum_gdpPercap = min(gdpPercap),
                         median_gdpPercap = median(gdpPercap),
                         average_gdpPercap = mean(gdpPercap),
                         standard_deviation_gdpPercap = sd(gdpPercap)))
```

#### Figure 1: Distribution of GDP per capita within each continent using histograms.
```{r}
ggplot(gapminder, aes(x=gdpPercap)) +
  facet_wrap(~ continent) +
  geom_histogram(aes(fill=continent), bins=20) +
  scale_x_log10("GDP per capita") +
  theme_bw() +
  labs(title="Distribution of GDP per capita") +
  scale_y_continuous("Frequency") +
  scale_color_discrete("Continent") +
  theme(legend.position="bottom",
        plot.title=element_text(hjust=0.5),
        strip.background = element_rect(fill="lightcyan"),
        strip.text = element_text(size=12, face="bold"))
```

#### Figure 2: Box plot of GDP per capita distribution.
```{r}
ggplot(gapminder, aes(x=continent, y=gdpPercap)) +
  geom_boxplot(aes(colour=continent), outlier.shape=1) +
  theme_bw() +
  labs(title="Distribution of GDP per capita") +
  scale_x_discrete("Continent") +
  scale_y_continuous("GDP per capita") +
  scale_color_discrete("Continent") +
  theme(plot.title=element_text(hjust=0.5))
```

#### Figure 3: Violin plot of GDP per capita distribution.
```{r}
ggplot(gapminder, aes(x=continent, y=gdpPercap)) +
  geom_violin(scale = "area", aes(fill=continent)) +
  theme_bw() +
  labs(title="Distribution of GDP per capita") +
  scale_x_discrete("Continent") +
  scale_y_continuous("GDP per capita") +
  scale_color_discrete("Continent") +
  theme(plot.title=element_text(hjust=0.5))
```

#### Figure 4: Distribution of GDP per capita within each continent using frequency plots.
```{r}
ggplot(gapminder, aes(x=gdpPercap)) +
  facet_grid(~ continent) +
  geom_freqpoly(aes(colour=continent), bins=20) +
  scale_x_log10("GDP per capita") +
  theme_bw() +
  labs(title="Distribution of GDP per capita") +
  scale_y_continuous("Frequency") +
  scale_color_discrete("Continent") +
  theme(legend.position = "bottom",
        plot.title=element_text(hjust=0.5),
        strip.background = element_rect(fill="pink"),
        strip.text = element_text(size=12, face="bold"))
```

## Task 3: Compute a trimmed mean of life expectancy for different years.

#### Table 1: 20% Trimmed mean of life expectancy for different years.

```{r results= 'asis'}
knitr::kable(gapminder %>%
               group_by(year) %>%
               summarize(trimmed_mean=round(mean(lifeExp, trim=0.20, na.rm=FALSE), digits=1)))
```

#### Figure 1: 20% Trimmed mean life expectancy for different years.

```{r}
ggplot(gapminder %>%
         group_by(year) %>%
         summarize(trimmed_mean=mean(lifeExp, trim=0.20, na.rm=FALSE)),
       aes(x=year, y=trimmed_mean)) +
  geom_point() +
  geom_smooth(se=FALSE, method="loess") +
  theme_bw() +
  labs(title="Trimmed Mean Life Expectancy by Year") +
  scale_x_continuous("Year") +
  scale_y_continuous("Trimmed Mean Life Expectancy") +
  theme(plot.title=element_text(hjust=0.5))
```

## Task 4: How is life expectancy changing over time on different continents?

#### Table 1: Maximum, minimum, and trimmed mean life expectancy for each continent over time.
```{r results= 'asis'}
knitr::kable(gapminder %>%
               select(continent, year, lifeExp) %>% 
               group_by(continent, year) %>%
               summarize(max_lifeExp=max(lifeExp),
                         min_lifeExp=min(lifeExp),
                         trimmed_mean=round(mean(lifeExp, trim=0.20, na.rm=FALSE), digit=1)))
```


#### Figure 1: Trends in life expectancy
```{r}
ggplot(gapminder, aes(x=year, y=lifeExp)) +
  geom_point(alpha=0.2, aes(colour=continent)) +
  geom_smooth(se=FALSE, method="loess", aes(colour=continent)) +
  theme_bw() +
  labs(title="Trends in Life Expectancy") +
  scale_x_continuous("Year") +
  scale_y_continuous("Life Expectancy") +
  scale_color_discrete("Continent") +
  theme(legend.position="top",
        plot.title=element_text(hjust=0.5))
```


#### Figure 2: Trends in life expectancy. 
```{r}
ggplot(gapminder, aes(x=year, y=lifeExp, group=country)) +
  facet_wrap(~ continent) +
  geom_line(alpha=0.3, aes(colour=continent)) +
  theme_bw() +
  labs(title="Trends in Life Expectancy") +
  scale_x_continuous("Year") +
  scale_y_continuous("Life Expectancy") +
  scale_color_discrete("Continent") +
  theme(legend.position = "bottom",
        plot.title=element_text(hjust=0.5),
        strip.background = element_rect(fill="peachpuff"),
        strip.text = element_text(size=12, face="bold"),
        axis.text.x=element_text(angle=90))
```


## Task 5: Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.

#### Table 1: Number of countries below mean life expectancy.
Mean life expectancy
```{r}
mean(gapminder$lifeExp)
```

```{r results= 'asis'}
knitr::kable(gapminder %>%
               select(country, continent, year, lifeExp) %>%
               filter(lifeExp < mean(lifeExp)) %>%
               group_by(continent, year) %>%
               summarize(countries_below_mean_lifeExp=n_distinct(country)))
```

#### Figure 1: Number of countries below mean life expectancy.
```{r}
low_lifeExp <- gapminder %>%
  select(country, continent, year, lifeExp) %>%
  filter(lifeExp < mean(lifeExp)) %>%
  group_by(continent, year) %>%
  summarize(countries_below_mean_lifeExp=n_distinct(country))

ggplot(low_lifeExp, aes(x=year, y=countries_below_mean_lifeExp)) +
  geom_line(aes(colour=continent)) +
  facet_wrap(~ continent) +
  theme_bw() +
  labs(title="Number of countries below mean life expectancy") +
  scale_x_continuous("Year") +
  scale_y_continuous("Number of countries") +
  scale_color_discrete("Continent") +
  theme(legend.position = "bottom",
        plot.title=element_text(hjust=0.5),
        strip.background = element_rect(fill="lightskyblue1"),
        strip.text = element_text(size=12, face="bold"))
```

## Task 6: Find countries with interesting stories! Which countries show the greatest change in life expectancy from 1990 to 2000?

#### Table 1: Top 10 countries showing the greatest positive change in life expectancy from 1992 to 2002.
```{r results= 'asis'}
pos_change <- gapminder %>%
               filter(year == 1992 | year == 2002) %>% 
               group_by(country) %>%
               arrange(year) %>% 
               mutate(lifeExp_change=lifeExp[year==2002] - lifeExp[year==1992]) %>% 
               arrange(desc(lifeExp_change)) %>%
               filter(year == 2002) %>% 
               select(country, continent, lifeExp_change) %>% 
               head(10)

knitr::kable(pos_change)
              
```

#### Table 2: Top 10 countries showing the greatest negative change in life expectancy from 1992 to 2002.

```{r results= 'asis'}
neg_change <- (gapminder %>%
                 filter(year == 1992 | year == 2002) %>% 
                 group_by(country) %>%
                 arrange(year) %>% 
                 mutate(lifeExp_change=lifeExp[year==2002] - lifeExp[year==1992]) %>% 
                 arrange(lifeExp_change)) %>%
                 filter(year == 2002) %>% 
                 select(country, continent, lifeExp_change) %>% 
                 head(10)

knitr::kable(neg_change)
```

#### Figure 1: Positive changes in life expectancy from 1992 to 2002
```{r}
ggplot(pos_change, aes(x=country, y=lifeExp_change)) +
  geom_col(aes(fill=continent)) +
  theme_bw() +
  labs(title="Changes in life expectancy from 1992 to 2002") +
  scale_x_discrete("Country") +
  scale_y_continuous("Change in life expectancy") +
  theme(legend.position="top",
        plot.title=element_text(hjust=0.5))
```

#### Figure 2: Negative changes in life expectancy from 1992 to 2002
```{r}
ggplot(neg_change, aes(x=country, y=lifeExp_change)) +
  geom_col(aes(fill=continent)) +
  theme_bw() +
  labs(title="Changes in life expectancy from 1992 to 2002") +
  scale_x_discrete("Country") +
  scale_y_continuous("Change in life expectancy") +
  theme(legend.position="top",
        plot.title=element_text(hjust=0.5),
        axis.text.x=element_text(angle=45))
```

